### 2010-08-25 Thesis/Chapter2.tex
- Update 25 August -
Started chapter5 (TLED description) + update of labels + package left indices + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@165 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{Description of motion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. Deformation implies the change in shape and\slash or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.

@@

	\subsection{Green strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-St.\,Venant strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green strain tensor is a symmetric second-order tensor. Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right]. \label{chap2:GreenTensor}
\end{align}

@@

	\subsection{Elasticity}	
If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}. The work done by the stress is generally dependent on the deformation path. However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation. It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.

@@

	\subsubsection*{Neo-Hookean}			
The neo-Hookean model was proposed by Ronald Rivlin in 1948. It can be derived from the Mooney-Rivlin model in the special case where $C_2 = 0$ in the strain energy density function. Therefore, we have:
\begin{equation}
U_0 = C_1 (I_B-3).
\end{equation}
and the constitutive equation for the stress may be written as:
\begin{equation}
\sigma = -p \mathbf{I} + 2 \rho C_1 \mathbf{B}.
\end{equation}
The neo-Hookean model provides a reasonable approximation of rubber-like material for moderate strains. It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.

>>>>

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. Deformation implies the change in shape and\slash or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.

@@

	\subsection{Green-Lagrange strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-Lagrange strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green-Lagrange strain tensor is a symmetric second-order tensor. Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right]. \label{chap2:GreenTensor}
\end{align}

@@

	\subsection{Elasticity}	\label{chap2:elasticity}
If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}. The work done by the stress is generally dependent on the deformation path. However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation. It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.

@@

	\subsubsection*{Neo-Hookean}			
The neo-Hookean model was proposed by Ronald Rivlin in 1948. It can be derived from the Mooney-Rivlin model in the special case where $C_2 = 0$ in the strain energy density function. Therefore, we have:
\begin{equation}
U_0 = C_1 (I_B-3).
\end{equation}
and the constitutive equation for the stress may be written as:
\begin{equation}
\sigma = -p \mathbf{I} + 2 \rho C_1 \mathbf{B}.
\end{equation}
The neo-Hookean model provides a reasonable approximation of rubber-like material for moderate strains. \TODO{Do I keep this range of validity, which seems somewhat limited...} It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.

### 2010-08-25 Thesis/Chapter3.tex
- Update 25 August -
Started chapter5 (TLED description) + update of labels + package left indices + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@165 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

		\subsubsection*{Mass matrix}
The mass matrix $ \mathbf{M} $ for the whole domain may be computed by summing the mass contributions from every elements:
\begin{equation}
\mathbf{M} = \sum_e \mathbf{M}^e,
\end{equation}
where $ \mathbf{M}^e $ is the mass of element $ \Omega_e $ given by:
\begin{equation}
\mathbf{M}^e = \int_{V^e} \rho^e \mathbf{H}^T \mathbf{H} dV
\end{equation}
where $ \rho $ the mass density, $ V^e $ the volume of the element and $ \mathbf{H} $ is the shape function matrix defined by \eqref{chap3:H}. This integral may be evaluated using a procedure of numerical integration and generally results in a band mass matrix (non-zero entries are confined to a diagonal band, comprising the main diagonal and more diagonals on either side). This method for computing the mass matrix is called \emph{variational mass lumping}.

@@

		\subsubsection*{Damping matrix}
In a similar way to the mass matrix, a damping matrix may be defined by summing the damping contributions from every elements:
\begin{equation}
\mathbf{D} = \sum_e \mathbf{D}^e,
\end{equation}
where $ \mathbf{D}^e $ is the damping of element $ \Omega_e $ given by:
\begin{equation}
\mathbf{D}^e = \int_{V^e} \kappa^e \mathbf{H}^T \mathbf{H} dV
\end{equation}
where $ \kappa^e $ is a damping parameter. In practice, the components of damping matrices are difficult to determine experimentally. Therefore, it is common to employ the \emph{Rayleigh damping}. The Rayleigh damping assumes that the damping may be written in the form of:
\begin{equation}
\mathbf{D}^e = \alpha \mathbf{M}^e + \beta \mathbf{K}^e,
\end{equation}
where $ \alpha $ and $ \beta $ are coefficients defining mass and stiffness proportional damping components. However, for systems with large degrees of freedom, it is difficult to guess meaningful values of $ \alpha $ and $ \beta $. \cite{Chowdhury03} have proposed a method to calculate these coefficients. In practice, further assumptions may only lead to include the mass proportional component ($ \beta = 0 $), which yields a diagonal damping matrix if a diagonal mass matrix is used.

@@

		\subsubsection*{The central difference method}	
The central difference method is one of the most popular explicit methods in computational mechanics. This method is based on central difference formulas for the velocity and acceleration. Let us the time $ t $ of the simulation be subdivided into time steps $ \Delta t $. Let us denote by $ \mathbf{U}^t $, $ \mathbf{\dot U}^t $ and $ \mathbf{\ddot U }^t $ the matrices of nodal displacements, velocities and accelerations, respectively, at time $ t $. The central difference method makes use of the following finite differences approximations for acceleration and velocity:
\begin{align}
\mathbf{\ddot U}^t &= \dfrac{1}{(\Delta t)^2} \left( \mathbf{U}^{t-\Delta t} - 2 \mathbf{U}^{t} +\mathbf{U}^{t+\Delta t} \right) \label{chap3:accelerationApprox}\\
\mathbf{\dot U}^t &= \dfrac{1}{2 \Delta t} \left( \mathbf{U}^{t+\Delta t} - \mathbf{U}^{t-\Delta t} \right). \label{chap3:velocityApprox}
\end{align}
One of the key advantage of the method is that the stifness term in \eqref{chap3:dynamic2} may be computed from
\begin{equation}
\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{F}(\mathbf{U}) = \sum_e \tilde{\mathbf{F}}^e,
\end{equation}
where $ \tilde{\mathbf{F}}^e $ are the global nodal force contributions due to stresses in element $ \Omega_e $. $ \tilde{\mathbf{F}} $ may be computed using strain-displacement equations, the constitutive equation and the element connectivity. Therefore, let us assume that all nodal forces $\mathbf{F}^t $  at time $ t $ are known.

By subtituting \eqref{chap3:accelerationApprox} and \eqref{chap3:velocityApprox} into \eqref{chap3:dynamic2} we obtain:
\begin{equation}
\left( \dfrac{\mathbf{M}}{(\Delta t)^2} + \dfrac{\mathbf{D}}{2 \Delta t} \right) \mathbf{U}^{t+\Delta t} = \mathbf{R}^t - \mathbf{F}^t + \dfrac{2 \mathbf{M}}{(\Delta t)^2} \mathbf{U}^t + \left( \dfrac{\mathbf{D}}{2 \Delta t} - \dfrac{\mathbf{M}}{(\Delta t)^2} \right) \mathbf{U}^{t-\Delta t}.
\end{equation}
Both the right-hand side of the equation and the factor in front $ \mathbf{U}^{t+\Delta t} $ may be evaluated, which gives the nodal displacements for the next time step. The entire update can be accomplished without solving any system of equations provided that the mass matrix $ \mathbf{M} $ and the damping matrix $ \mathbf{D} $ are diagonal. As we have seen in section \ref{chap3:wordOnMatrices}, these approximations are often used in practice. And this is the characteristic of an explicit method: the time integration of the general system of equations for finite element model does not require the solution of any equations. This is a key point because it gives explicit methods a tremendous computational advantage. However, explicit methods are only conditionally stable. If the time step exceeds a critical value, the solution may be erroneous. This critical time step depends on the mesh and the material properties. It will decrease with mesh refinement and increasing stiffness of the material.

@@

>>>>

		\subsubsection*{Mass matrix}
The mass matrix $ \mathbf{M} $ for the whole domain may be computed by summing the mass contributions from every elements:
\begin{equation}
\mathbf{M} = \sum_e \mathbf{M}^e,
\end{equation}
where $ \mathbf{M}^e $ is the mass of element $ \Omega_e $ given by:
\begin{equation}
\mathbf{M}^e = \int_{V^e} \rho^e \mathbf{H}^T \mathbf{H} \, dV
\end{equation}
where $ \rho $ the mass density, $ V^e $ the volume of the element and $ \mathbf{H} $ is the shape function matrix defined by \eqref{chap3:H}. This integral may be evaluated using a procedure of numerical integration and generally results in a band mass matrix (non-zero entries are confined to a diagonal band, comprising the main diagonal and more diagonals on either side). This method for computing the mass matrix is called \emph{variational mass lumping}.

@@

		\subsubsection*{Damping matrix}
In a similar way to the mass matrix, a damping matrix may be defined by summing the damping contributions from every elements:
\begin{equation}
\mathbf{D} = \sum_e \mathbf{D}^e,
\end{equation}
where $ \mathbf{D}^e $ is the damping of element $ \Omega_e $ given by:
\begin{equation}
\mathbf{D}^e = \int_{V^e} \kappa^e \mathbf{H}^T \mathbf{H} \, dV
\end{equation}
where $ \kappa^e $ is a damping parameter. In practice, the components of damping matrices are difficult to determine experimentally. Therefore, it is common to employ the \emph{Rayleigh damping}. The Rayleigh damping assumes that the damping may be written in the form of:
\begin{equation}
\mathbf{D}^e = \alpha \mathbf{M}^e + \beta \mathbf{K}^e,
\end{equation}
where $ \alpha $ and $ \beta $ are coefficients defining mass and stiffness proportional damping components. However, for systems with large degrees of freedom, it is difficult to guess meaningful values of $ \alpha $ and $ \beta $. \cite{Chowdhury03} have proposed a method to calculate these coefficients. In practice, further assumptions may only lead to include the mass proportional component ($ \beta = 0 $), which yields a diagonal damping matrix if a diagonal mass matrix is used.

@@

		\subsubsection*{The central difference method}		\label{chap3:centralDifferenceMethod}
The central difference method is one of the most popular explicit methods in computational mechanics. This method is based on central difference formulas for the velocity and acceleration. Let us the time $ t $ of the simulation be subdivided into time steps $ \Delta t $. Let us denote by $ \mathbf{U}^t $, $ \mathbf{\dot U}^t $ and $ \mathbf{\ddot U }^t $ the matrices of nodal displacements, velocities and accelerations, respectively, at time $ t $. The central difference method makes use of the following finite differences approximations for acceleration and velocity:
\begin{align}
\mathbf{\ddot U}^t &= \dfrac{1}{(\Delta t)^2} \left( \mathbf{U}^{t-\Delta t} - 2 \mathbf{U}^{t} +\mathbf{U}^{t+\Delta t} \right) \label{chap3:accelerationApprox}\\
\mathbf{\dot U}^t &= \dfrac{1}{2 \Delta t} \left( \mathbf{U}^{t+\Delta t} - \mathbf{U}^{t-\Delta t} \right). \label{chap3:velocityApprox}
\end{align}
One of the key advantage of the method is that the stifness term in \eqref{chap3:dynamic2} may be computed from
\begin{equation}
\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{F}(\mathbf{U}) = \sum_e \tilde{\mathbf{F}}^e,
\end{equation}
where $ \tilde{\mathbf{F}}^e $ are the global nodal force contributions due to stresses in element $ \Omega_e $. $ \tilde{\mathbf{F}} $ may be computed using strain-displacement equations, the constitutive equation and the element connectivity. Therefore, let us assume that all nodal forces $\mathbf{F}^t $  at time $ t $ are known.

By subtituting \eqref{chap3:accelerationApprox} and \eqref{chap3:velocityApprox} into \eqref{chap3:dynamic2} we obtain:
\begin{equation}
\label{chap3:relationCDM}
\left( \dfrac{\mathbf{M}}{(\Delta t)^2} + \dfrac{\mathbf{D}}{2 \Delta t} \right) \mathbf{U}^{t+\Delta t} = \mathbf{R}^t - \mathbf{F}^t + \dfrac{2 \mathbf{M}}{(\Delta t)^2} \mathbf{U}^t + \left( \dfrac{\mathbf{D}}{2 \Delta t} - \dfrac{\mathbf{M}}{(\Delta t)^2} \right) \mathbf{U}^{t-\Delta t}.
\end{equation}
Both the right-hand side of the equation and the factor in front $ \mathbf{U}^{t+\Delta t} $ may be evaluated, which gives the nodal displacements for the next time step. The entire update can be accomplished without solving any system of equations provided that the mass matrix $ \mathbf{M} $ and the damping matrix $ \mathbf{D} $ are diagonal. As we have seen in section \ref{chap3:wordOnMatrices}, these approximations are often used in practice. And this is the characteristic of an explicit method: the time integration of the general system of equations for finite element model does not require the solution of any equations. This is a key point because it gives explicit methods a tremendous computational advantage. However, explicit methods are only conditionally stable. If the time step exceeds a critical value, the solution may be erroneous. This critical time step depends on the mesh and the material properties. It will decrease with mesh refinement and increasing stiffness of the material.

@@

\paragraph*{\TODO{Cholesky decomposition?}}
